
# A Brief R Tutorial

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="/Users/Wen/百度云同步盘/01_Teaching/01_My_Courses/11_QuantMethods_I/")
```

## Essential Basics - intro

In R, every method is a function: `name` + `()`. For example, in your R console, try type `help()`. 


## Atomic Data Types - atomic-dtype

There are four commonly used atomic data types. 

- logical: e.g. TRUE, FALSE
- integer: e.g. -2, 0, 1, 2
- numeric: e.g., -2.0, 0.05, 1.2
- character: e.g., "-2.0", "cat", "dog"

### Logical - logical

Logical is simply `TRUE` or `FALSE`. To note, `T` is a shorthand for `TRUE`. They are strictly identical. 

```{r}
TRUE
T
```

To check the data type of `TRUE`, use the `class()` function.
```{r}
class(TRUE)
class(T)
```

To note, you **MUST NOT** use `True` or `False`, which would result in an error. 
```{r, error=TRUE}
True
```


### Integer - integer

If you directly type `2` into the R console, you might think it's an integer, but it is in fact of numeric data type. To note, numeric data type is also called **floating point** （浮点型数据） in other programming languages. 

```{r}
2; class(2)
```

The `;` sign is used to separate commands such that two lines can be written succinctly（简洁的） in one line. In our case, ```2; class(2)``` is identical to 

```{r, eval=FALSE}
2
class(2)
```

To have an integer, you can attach an `L` to the integer. 
```{r}
2L; class(2L)
```

```{r}
x = dim(swiss)
x[1]
```


```{r}
xdat = read.csv("Data/demo_data.csv", stringsAsFactors = F)
xvec = xdat[,"session"]  # extract the feature 'session'
xvec = xdat$session       # identical
cnts = table(xvec)
cnts
pcts = 100 * table(xvec) / sum( table(xvec) )
pcts
mat = cbind(cnts, pcts)
mat
```


